Precise spiking motifs in neurobiological and neuromorphic data

### 2022-12-19 manuscript.tex
Update on Overleaf.

There are different solutions to provide with a distance between two given spike trains. A known measure is the Victor-Purpura distance, which overcomes inconsistencies experienced with a \remove[JN]{simple} firing rate (Poisson model) of spike trains~\citep{victor_nature_1996}. \change[JN]{In a later study,van_rossum_novel_2001 have defined a distance that uses a time constant as a parameter}{Then a study tries to solve the problem by including a time constant as a parameter}~\citet{van_rossum_novel_2001}. \change[JN]{Depending on this parameter, the distance interpolates between a coincidence detector and a rate difference counter.}{This parameter will then be used to interpolate the distance between a coincidence detector and a rate difference counter.} Such distances were extended to non-Euclidean metrics and use morphological manipulations to compute spike train dissimilarity~\citep{kreuz_measuring_2007}. Mathematically, the stability of distance measures induced by level-crossing sampling can be evaluated~\citep{moser_stability_2014}, notably in light of the so-called Weyl's discrepancy measure~\citep{weyl_ueber_1916}, which may lead to the definition of a cross-correlation measure \change[JN]{which}{. An interesting conclusion since cross-correlation measure is} which is adapted to the event-based nature of spiking signals. These observations lead to the intuition that each distance may be as well  as the optimal solution of a generative model for these measures, possibly through non-linear relations~\citep{aronov_non-euclidean_2004}. % ~\citep{}

%Let us now focus on distances defined based on a generative model formed by using precise spiking motifs. 
Concerning spike timings,~\citet{levakova_review_2015} \change[Antoine]{has}{have} \change[JN]{reviewed the methods for neuronal response latency estimation exists and includes Bayesian binning}{reviewed existing methods for estimating the latency of neural responses that include Bayesian binning.}. \change[JN]{More broadly }{Alternatively }, unitary event analysis can be performed by a statistical model of coincidence detection~\citep{grun_unitary_2002-1}. This was extensively used in detecting above chance significant synchronous patterns~\citep{grun_unitary_2010}, in particular in recordings of pairs of neurons (see~\citep{riehle_spike_1997} for instance), and has been extended to non-stationary data~\citep{grun_unitary_2002}. A method to detect significant patterns of synchronous spiking in a subset of massively parallel spike trains in the presence of background activity can be defined using the statistical evaluation of synchronous spike patterns extracted by frequent item set mining~\citep{torre_statistical_2013}. By the same group, the SPADE, CAD or ASSET algorithms are methods for identification of spike patterns in massively parallel spike trains (the spiking activity of tens to hundred(s) of neurons recorded in parallel) by identifying fine temporal correlations in the$~\ms$ precision range~\citep{quaglio_methods_2018}. This was recently extended in~\citep{stella_3d-spade_2019} in order to find re-occurring patterns in parallel spike train data, and to determine their statistical significance. The extension improves the performance in the presence of patterns with different durations, as demonstrated by application to various synthetic data \change[LP]{}{such as the synthetic data for synfire chains (see Figure~\ref{fig:Diesmann_et_al_1999_spade}),} such as surrogates generated to evaluate precisely timed higher-order spike correlations~\citep{stella_comparing_2022}.

@@

However, most biological neural systems are obviously more efficient than current state-of-the-art vision systems, both in terms of efficiency (accuracy), in speed (latency), and energy consumption.}{So far, this approach does not outperform classical architectures both in term of training efficiency and performances}~\citep{davies_loihi_2018}\add[Antoine]{. However, they remain the best candidates to reproduce biological neural systems and their capacities both in terms of accuracy, speed and energy consumption.} There is therefore an immense gap in the way we understand biology to translate it to the efficiency of SNNs. To go beyond the state-of-the-art, we will focus here on one core computation of a spiking neuron, that is, its ability to switch from the classical integrator mode (summing analog currents on its \change[Antoine]{synapses}{dendrites}) to a synchrony detector where it emits a spike whenever presynaptic spiking inputs are synchronized~\citep{abeles_role_1982,paugam-moisy_computing_2012}. In particular, we will explore different existing architectures which are able to overcome the diversity of input presynaptic patterns and learn to detect stable spiking motifs, that is, volleys of spikes which are stable up to a certain onset time (see Figure~\ref{fig:THC}). These models will be compared in light of neuroscientific and computational perspectives. %We review in this section theoretical and computational foundations of spiking motif detection. %In particular, our objective is to describe methods which fully exploit the capacity of spiking neurons to detect synchronous patterns.

@@

As we saw, most SNN, and in particular those adapted from analogous deep-learning-like architectures, rely on an encoding of information based on a continuously varying firing-rate. A notable exception of a SNN using precise spike timings is the \textit{polychronization} model of~\citet{izhikevich_polychronization_2006}. This model is based on a random recurrent model of spiking neurons including synaptic delays chosen randomly from a range of biologically realistic delays (from $0$ to $20~\ms$) and whose weights evolved with a Spike-Time Dependent Plasticity (STDP) learning rule~\citep{markram_regulation_1997}. \change[Antoine]{It is consensus that spike timing (STDP) plays a crucial role in the development of synaptic efficacy for many different kinds of neurons}{It was shown that spike timing (STDP) has an impact on the development of synaptic efficacy for many kinds of neurons}~\citep{caporale_spike_2008}. Delays are defined as the total time taken for a spike to be conducted from one presynaptic neuron's soma to the efferent postsynaptic neuron's soma. It is worth mentioning that only the weights are changed using the STDP rule and that the set of delays is set randomly at initialization and that delays are then ``frozen'' for the rest of the simulation. Due to the interplay between the delays and STDP, the spiking neurons spontaneously self-organize into groups and generate patterns of stereotypical polychronous activity, i.e.~exhibit reproducible time-locked firing patterns which  the author defined as ``polychronous groups'' (PGs). One core ingredient of this model is the fact that the neurons composing a group fire at different times, but due to the heterogeneous delays, the spikes reach the postsynaptic neuron at the same time. This synchrony of arrival at the soma of the neuron leads to the summation of the excitatory post-synaptic potentials evoked by each spike, and thus to the crossing of the voltage threshold and to the discharge of a spike (see Figure~\ref{fig:izhikevich}). According to the STDP rule, the group of neurons involved in this polychronous activity will see their synaptic weight increase and, thus, may consolidate the formation of a polychronous group. 

>>>>

There are different solutions to provide with a distance between two given spike trains. A known measure is the Victor-Purpura distance, which overcomes inconsistencies experienced with a \remove[JN]{simple} firing rate (Poisson model) of spike trains~\citep{victor_nature_1996}. \change[JN]{In a later study, van_rossum_novel_2001 have defined a distance that uses a time constant as a parameter}{Then a study tries to solve the problem by including a time constant as a parameter}~\citet{van_rossum_novel_2001}. \change[JN]{Depending on this parameter, the distance interpolates between a coincidence detector and a rate difference counter.}{This parameter will then be used to interpolate the distance between a coincidence detector and a rate difference counter.} Such distances were extended to non-Euclidean metrics and use morphological manipulations to compute spike train dissimilarity~\citep{kreuz_measuring_2007}. Mathematically, the stability of distance measures induced by level-crossing sampling can be evaluated~\citep{moser_stability_2014}, notably in light of the so-called Weyl's discrepancy measure~\citep{weyl_ueber_1916}, which may lead to the definition of a cross-correlation measure \change[JN]{which}{. An interesting conclusion since cross-correlation measure is} which is adapted to the event-based nature of spiking signals. These observations lead to the intuition that each distance may be as well  as the optimal solution of a generative model for these measures, possibly through non-linear relations~\citep{aronov_non-euclidean_2004}. % ~\citep{}

%Let us now focus on distances defined based on a generative model formed by using precise spiking motifs. 
Concerning spike timings,~\citet{levakova_review_2015} \change[Antoine]{has}{have} \change[JN]{reviewed the methods for neuronal response latency estimation exists and includes Bayesian binning}{reviewed existing methods for estimating the latency of neural responses that include Bayesian binning.}. \change[JN]{More broadly }{Alternatively }, unitary event analysis can be performed by a statistical model of coincidence detection~\citep{grun_unitary_2002-1}. This was extensively used in detecting above chance significant synchronous patterns~\citep{grun_unitary_2010}, in particular in recordings of pairs of neurons (see~\citep{riehle_spike_1997} for instance), and has been extended to non-stationary data~\citep{grun_unitary_2002}. A method to detect significant patterns of synchronous spiking in a subset of massively parallel spike trains in the presence of background activity can be defined using the statistical evaluation of synchronous spike patterns extracted by frequent item set mining~\citep{torre_statistical_2013}. By the same group, the SPADE, CAD or ASSET algorithms are methods for identification of spike patterns in massively parallel spike trains (the spiking activity of tens to hundred(s) of neurons recorded in parallel) by identifying fine temporal correlations in the$~\ms$ precision range~\citep{quaglio_methods_2018}. This was recently extended in~\citep{stella_3d-spade_2019} in order to find re-occurring patterns in parallel spike train data, and to determine their statistical significance. The extension improves the performance in the presence of patterns with different durations, as demonstrated by application to various synthetic data \add[LP]{such as the synthetic data for synfire chains (see Figure}~\ref{fig:Diesmann_et_al_1999_spade}), such as surrogates generated to evaluate precisely timed higher-order spike correlations~\citep{stella_comparing_2022}.

@@

However, most biological neural systems are obviously more efficient than current state-of-the-art vision systems, both in terms of efficiency (accuracy), in speed (latency), and energy consumption.}{So far, this approach does not outperform classical architectures both in term of training efficiency and performances}~\citep{davies_loihi_2018}\add[Antoine]{. However, they remain the best candidates to reproduce biological neural systems and their capacities both in terms of accuracy, speed and energy consumption.} There is therefore an immense gap in the way we understand biology to translate it to the efficiency of SNNs. To go beyond the state-of-the-art, we will focus here on one core computation of a spiking neuron, that is, its ability to switch from the classical integrator mode (summing analog currents on its \change[Antoine]{synapses}{dendrites}) \change[Antoine]{to a synchrony detector where it emits a spike whenever presynaptic spiking inputs are synchronized}{to a detector of precise spiking motifs}~\citep{paugam-moisy_computing_2012}. In particular, we will explore different existing architectures which are able to overcome the diversity of input presynaptic patterns and learn to detect stable spiking motifs, that is, volleys of spikes which are stable up to a certain onset time (see Figure~\ref{fig:THC}). These models will be compared in light of neuroscientific and computational perspectives. %We review in this section theoretical and computational foundations of spiking motif detection. %In particular, our objective is to describe methods which fully exploit the capacity of spiking neurons to detect synchronous patterns.

@@

As we saw, most SNN, and in particular those adapted from analogous deep-learning-like architectures, rely on an encoding of information based on a continuously varying firing-rate. Notable exceptions of a SNN using precise spike timings are the \add[Antoine]{time encoding machine by}~\cite{lazar_time_2004} and \textit{polychronization} model of~\citet{izhikevich_polychronization_2006}. \add[Antoine]{In this section we focus on the polychronization model} based on a random recurrent model of spiking neurons including synaptic delays chosen from a range of biologically realistic delays (from $0$ to $20~\ms$) and whose weights evolved with a Spike-Time Dependent Plasticity (STDP) learning rule~\citep{markram_regulation_1997}. \change[Antoine]{It is consensus that spike timing (STDP) plays a crucial role in the development of synaptic efficacy for many different kinds of neurons}{It was shown that spike timing (STDP) has an impact on the development of synaptic efficacy for many kinds of neurons}~\citep{caporale_spike_2008}. Delays are defined as the total time taken for a spike to be conducted from one presynaptic neuron's soma to the efferent postsynaptic neuron's soma. It is worth mentioning that only the weights are changed using the STDP rule and that the set of delays is set randomly at initialization and that delays are then ``frozen'' for the rest of the simulation. Due to the interplay between the delays and STDP, the spiking neurons spontaneously self-organize into groups and generate patterns of stereotypical polychronous activity, i.e.~exhibit reproducible time-locked firing patterns which  the author defined as ``polychronous groups'' (PGs). One core ingredient of this model is the fact that the neurons composing a group fire at different times, but due to the heterogeneous delays, the spikes reach the postsynaptic neuron at the same time. This synchrony of arrival at the soma of the neuron leads to the summation of the excitatory post-synaptic potentials evoked by each spike, and thus to the crossing of the voltage threshold and to the discharge of a spike (see Figure~\ref{fig:izhikevich}). According to the STDP rule, the group of neurons involved in this polychronous activity will see their synaptic weight increase and, thus, may consolidate the formation of a polychronous group. 
